\thetitle

### 2021-03-03 writing/challenges.tex
intro, landscape, why-rational: rewording
To clarify parts that confused me

At first glance implementing the methodology of section~\ref{sec:why-rational}
seems straightforward. The rational programmer is a simple algorithm and testing
whether the rational programmer hypothesis holds amounts to running the
algorithm on a collection of scenarios.

@@

Thus, an experimental analysis of blame needs a mostly new set of mutators.
Roughly speaking, the new mutators inject type errors into fully typed programs.
Applying such a mutator to any typed component produces a mutant component.  A
debugging scenario results from removing the types from the mutant. For the
design of such mutators, the authors relied on their own extensive programming
experience though not without discovering a major pitfall: some of their
original mutators systematically produced programs that immediately revealed the
source of the impedance mismatch. About two thirds of the original mutators
survived this first scrutiny testing. All of the remaining ones yield {\em
interesting debugging scenarios\/} (see section~\ref{sub:mutate-interesting}).

>>>>

At first glance implementing the method of section~\ref{sec:why-rational}
seems straightforward. The rational programmer is a simple algorithm and testing
whether the rational programmer hypothesis holds amounts to running the
algorithm on a collection of scenarios.

@@

Thus, an experimental analysis of blame needs a mostly new set of mutators.
Roughly speaking, the new mutators inject type errors into fully typed programs.
Applying such a mutator to any typed component produces a mutant component.  A
debugging scenario results from removing the types from the mutant. For the
design of such mutators, the authors relied on their own extensive programming
experience though not without discovering a major pitfall: some of their
original mutators systematically produced programs that immediately revealed the
source of the impedance mismatch. About four fifths of the original mutators
survived this first scrutiny testing. All of the remaining ones yield {\em
interesting debugging scenarios\/} (see section~\ref{sub:mutate-interesting}).

### 2021-03-03 writing/introduction.tex
intro, landscape, why-rational: rewording
To clarify parts that confused me

Theoreticians of gradual typing have focused on blame theorems from the very
beginning~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed
[components]\footnote{The original authors got this word wrong.} can't be
blamed'' turned the theorem into a slogan~\cite{wf-esop-2009}. Academic systems
(Reticulated Python~\cite{vsc-dls-2019, vss-popl-2017, vksb-dls-2014} and Typed
Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010}) come
with sophisticated checking and blame assignment strategies (see sec.~\ref{sec:landscape}). Their academic creators embrace
the idea that blame can help practicing programmers find impedance mismatches
between the types they added to a software system and the behavior of the
remaining untyped components.

@@

This contrast between theory and applications of gradual typing raises the question 
\begin{quote}
 \it
 whether blame assignment adds any value to a gradually typed language,
 especially for the benefit of the working programmer.
\end{quote}
Given the long-standing academic interest in blame and its complete absence in
industrial systems, it comes as an even bigger surprise that the research
literature and the industrial blog world do not discuss any possible answers.
Instead, when language designers make relevant decisions, they seem to answer it
one way or another without any scientific justification. Then again, the
community has thus far failed to offer a method for evaluating blame assignment.

@@

The paper's {\it second contribution\/} is a set of {\em results from applying
the evaluation method to three distinct blame assignment strategies in
approximately 72,200 different scenarios\/} (see sec.~\ref{sec:results}): (1)~{\it
Transient\/}, i.e. Reticulated's tracking of typed/untyped boundaries~\cite{vss-popl-2017}; (2)~{\it
Natural\/}, i.e. the use of a higher-order contract system and its blame
assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e. the approach of
industrial systems, which forgo blame in favor of error messages from the safety
checks of the underlying language. The results (see
sec.~\ref{sec:discussion}) are at least somewhat surprising.  In principle
they validate the conjectures behind the work of theoreticians.  First, a good
blame assignment strategy helps with the search for impedance mismatches between
the specified types and the behavior of untyped components.  Second, Natural's
wrapper-based blame tracking is more useful than Transient's ``collective
blame'' tracking algorithm, which in turn is superior to Erasure. {\em But\/},
the application of the method also indicates problems with the expectations of
theoreticians. The first problem is that the cost of Transient's blame can be
huge. The second one concerns the difference between the methods; neither is
{\it Natural\/} vastly better than {\it Transient\/} nor are the two blame
assignment methods clearly preferable to {\it Erasure\/}.  In turn, these
problems suggests that, on one hand, the existing theory does not predict
practice properly, and on the other hand, the existing practice may need to
think of additional experiments to collect more data.

>>>>

Theoreticians of gradual typing have focused on blame theorems from the very
beginning~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed
[components]\footnote{The original authors got this word wrong.} can't be
blamed'' turned the theorem into a slogan~\cite{wf-esop-2009}. Academic systems
(Reticulated Python~\cite{vsc-dls-2019, vss-popl-2017, vksb-dls-2014} and Typed
Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010}) come
with sophisticated checking and blame assignment strategies (see sec.~\ref{sec:landscape}). Their academic creators embrace
the idea that blame can help practicing programmers find impedance mismatches: disagreements
between the types they added to a software system and its behavior.

@@

This contrast between theory and applications of gradual typing raises the question of
\begin{quote}
 \it
 whether blame assignment adds any value to a gradually typed language,
 especially for the benefit of the working programmer.
\end{quote}
Given the long-standing academic interest in blame and its complete absence in
industrial systems, it comes as an even bigger surprise that the research
literature and the industrial blog world do not discuss any possible answers.
Instead, when language designers make relevant decisions, they seem to answer it
one way or another without any scientific justification. Then again, the
community has thus far failed to offer a method for evaluating blame assignment.

@@

The paper's {\it second contribution\/} is a set of {\em results from applying
the evaluation method to three distinct blame assignment strategies in
approximately 72,200 different scenarios\/} (see sec.~\ref{sec:results}): (1)~{\it
Transient\/}, i.e. Reticulated's tracking of typed/untyped boundaries~\cite{vss-popl-2017}; (2)~{\it
Natural\/}, i.e. the use of a higher-order contract system and its blame
assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e. the approach of
industrial systems, which forgo blame in favor of error messages from the safety
checks of the underlying language. The results (see
sec.~\ref{sec:discussion}) are at least somewhat surprising.  In principle
they validate the conjectures behind the work of theoreticians.  First, a good
blame assignment strategy helps with the search for impedance mismatches between
the specified types and the behavior of untyped components.  Second, Natural's
wrapper-based blame tracking is more useful than Transient's ``collective
blame'' tracking algorithm, which in turn is superior to Erasure. {\em But\/},
the application of the method also indicates problems with the expectations of
theoreticians. The first problem is that the cost of Transient's blame can be
huge. The second one concerns the difference between the methods; neither is
{\it Natural\/} vastly better than {\it Transient\/} nor are the two blame
assignment methods clearly preferable to {\it Erasure\/}.  In turn, these
problems suggest that, on one hand, the existing theory does not predict
practice properly, and on the other hand, the existing practice may need
additional experiments to collect more data.

### 2021-03-03 writing/landscape.tex
intro, landscape, why-rational: rewording
To clarify parts that confused me

The three blame strategies rely on three different ways of catching problems
with types at run-time: {\em Natural\/}, {\em Transient\/}, and {\em
Erasure\/}. This section summarizes these options with one illustrative example
using Racket syntax. The informed reader may wish to merely scan the
self-contained section.

Consider the program sketch in figure~\ref{fig:example}. Each box represents a
module: the top bar lists the name and whether it is using typed (\typecolor) or
untyped (\dyncolor) syntax:
\begin{description}

\item[\texttt{pack-lib}] (at the top right) represents a library that provides,
among others, a function {\tt pack}. As the documentation says, this function
consumes JSON data and packages it in a dictionary. by mistake, though, the
function returns a hash table instead of the association list its documentation
promises.

@@

\item[\texttt{client}] (at the bottom right) uses {\tt pack}
indirectly. Specifically, it goes through the two intermediary typed modules to
import them. Imagine a programmer who relies on the types in the \typecolor\
modules as checked documentation but prototypes the client in an untyped manner.

\end{description}
Each of the three semantics assigns a different outcome to the client module and
signals different kinds of errors.

@@

As this analysis implies, if a return-type check fails, the problem is with
the untyped module that supplied an incorrect function, here {\tt pack-lib}. If
an argument-type check fails, responsibility lies with {\tt client}. Due to the
proxies wrapped around functions, Natural can easily track the responsible party
for each check. Thus, in the example of figure~\ref{fig:example}, as
\texttt{pack} returns, the return-type check fails, and Natural blames
\texttt{pack-lib}.

@@

In order to locate the responsible party for failed checks, Transient
maintains a map from values to the boundaries between typed/untyped modules that
they cross, plus the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{pack-lib} to {\tt crypto-pack-lib} and
from {\tt crypto-pack-lib} to {\tt client} with the type that appears in the
{\tt required/typed} forms in the example. Since the failed check corresponds to
the return type of \texttt{pack}, Transient blames the sources of the two
boundary crossings, which in this particular case is \texttt{pack-lib}. In
general though, Transient blames more than one party. In fact, the theoretical
work of~\citet{gfd-oopsla-2019} shows that for some programs Transient
constructs a blame set that excludes responsible parties and includes modules
irrelevant to the failing check.

Under {\it the Erasure semantics\/}, the compiler checks the specified types and
then discards them when it generates code. The generated code includes run-time
checks that ensure the dynamic safety of all operations as specified in the
underlying untyped language. Hence, neither the call to {\tt typed-pack} nor the
call to {\tt crypto-pack} signals an error due to the gradual type system. When,
at some later point, {\tt client} tries to inspect the elements of the lists
that \texttt{typed-pack} and {\tt crypto-pack} are supposed to produce, Racket's
safety check signal a violation and point to some place in {\tt client}. The
information in this exception, plus its stack trace, may help the programmer
find the source of the impedance mismatch between the specified return types of
{\tt pack} in the two typed modules and the actual return value of {\tt pack}.

>>>>

The three blame strategies rely on three different ways of catching problems
with types at run-time: {\em Natural\/}, {\em Transient\/}, and {\em
Erasure\/}. This self-contained section summarizes these options with one illustrative example
using Racket syntax. The informed reader may wish to merely scan it.

Consider the program sketch in figure~\ref{fig:example}. Each box represents a
module: the top bar lists the name and whether it is using typed (\typecolor) or
untyped (\dyncolor) syntax.
\begin{description}

\item[\texttt{pack-lib}] (at the top right) represents a library that provides,
among others, a function {\tt pack}. As the documentation says, this function
consumes JSON data and packages it in a dictionary. By mistake, though, the
function returns a hash table instead of the association list its documentation
promises.

@@

\item[\texttt{client}] (at the bottom right) uses {\tt pack}
indirectly. Specifically, it goes through the two intermediary typed modules to
use it. Imagine a programmer who relies on the types in the \typecolor\
modules as checked documentation but prototypes the client in an untyped manner.

\end{description}
Due to the mistake in {\tt pack-lib}, each of the three semantics assigns a
different outcome to the client module and signals different kinds of errors.

@@

As this analysis implies, if a return-type check fails, the problem is with
the untyped module that supplied an incorrect function, here {\tt pack-lib}. If
an argument-type check fails, responsibility lies with {\tt client}. Due to the
proxies wrapped around functions, Natural can easily track the responsible party
for each check. Thus, in the example of figure~\ref{fig:example}, as
\texttt{pack} returns, the return-type check fails and Natural blames
\texttt{pack-lib}.

@@

In order to locate the responsible party for failed checks, Transient
maintains a map from values to the boundaries between typed and untyped modules that
they cross, plus the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{pack-lib} to {\tt crypto-pack-lib} and
from {\tt crypto-pack-lib} to {\tt client} with the type that appears in the
{\tt required/typed} forms in the example. Since the failed check corresponds to
the return type of \texttt{pack}, Transient blames the sources of the two
boundary crossings, which in this particular case is \texttt{pack-lib}. In
general though, Transient blames more than one party. In fact, the theoretical
work of~\citet{gfd-oopsla-2019} shows that for some programs Transient
constructs a blame set that excludes responsible parties and includes modules
irrelevant to the failing check.

Under {\it the Erasure semantics\/}, the compiler checks the specified types and
then discards them when it generates code. The generated code includes run-time
checks that ensure the dynamic safety of all operations as specified in the
underlying untyped language. Hence, neither the call to {\tt typed-pack} nor the
call to {\tt crypto-pack} signals an error due to the gradual type system. If
at some later point {\tt client} tries to inspect the elements of the lists
that \texttt{typed-pack} and {\tt crypto-pack} are supposed to produce, Racket's
safety checks signal a violation and point to some place in {\tt client}. The
information in this exception, plus its stack trace, may help the programmer
find the source of the impedance mismatch between the specified return types of
{\tt pack} in the two typed modules and the actual return value of {\tt pack}.

### 2021-03-03 writing/why-rational-methodology.tex
intro, landscape, why-rational: rewording
To clarify parts that confused me

By contrast, when the rational programmer fails, it questions programming
language research. Specifically, it is indicative of the limited predictive
power of programming language theory with respect to the use of languages in
practice. Put differently, it suggests flaws in language design due to
misleading predictions.

@@

\item Does a language design provide information that guides the actions of a
 rational programmer?

@@

\item The error messages of blame-assignment mechanisms include suspect
violations at the boundary of typed and untyped code fragments. The
Wadler-Findler slogan suggests the problem is a lack of types, meaning the
addition of types to the untyped fragment is the most natural action.

@@

\item The theory theories implies that blame assignment constrains the search
space that a developer must inspect to find the problem.

\end{enumerate}
Sadly, seeing an algorithm is different from using the algorithm to conduct a
data-gathering experiments. The specific challenges are spelled out in the next
section, and the following three sections explains ways of overcoming them.

>>>>

By contrast, when the rational programmer fails, it questions programming
language research. Specifically, it is indicative of the limited predictive
power of programming language theory with respect to the use of languages in
practice. Put differently, it suggests flaws in language design revealed by
misleading predictions.

@@

\item Does a language design provide information that guides the actions of the
 rational programmer?

@@

\item The error messages of blame-assignment mechanisms include suspect
locations at the boundary of typed and untyped code fragments.
The Wadler-Findler slogan suggests that the source of the problem is concealed due to a lack of types, so
adding types to the untyped fragment should lead to the source of the impendance mismatch.

@@

\item The theory conjectures that blame assignment constrains the search
space that a developer must inspect to find the problem.

\end{enumerate}
Unfortunately, seeing an algorithm is different from using the algorithm to conduct
data-gathering experiments. The specific challenges are spelled out in the next
section, and the following three sections explains ways of overcoming them.

### 2021-03-03 writing/why-rational.tex
methodlogy
intro, landscape, why-rational: rewording
To clarify parts that confused me

This paper dubs Lazarek et. al's idea {\em the rational programmer\/}. Like Stuart
Mill's {\it homo economicus\/}, the rational programmer approximates the
behavior of a software developer who reduces time spent on a task by exploiting the
available information. In the context of gradual typing, the rational programmer
has two pieces of information when an impedance mismatch signals exceptional
behavior: the error message and the state of the program. Hence, the most
rational procedure is to use the former to improve the latter. Specifically,
a rational programmer translates the Wadler-Findler slogan into a debugging
method, searching for the source of the impedance mismatch by adding type
annotations to the untyped parts of the program identified in the error.
If the type checker rejects an annotation that is considered correct, the
rational programmer has found the source of the problem.  Measuring this simulated
programmer's behavior on a large number of debugging scenarios yields data that
is similar to data collected in a human-facing study.

@@

Like {\it homo economicus\/}, the rational programmer is an approximation.
People do not behave in a purely rational manner as economic actors, and they
also do not do so as software developers. The point is not to deny the existence
of ``lucky hunches'' programmers or ``tinkering works'' approaches and so on. It
is also not to deny that equipping entire modules with types represents an
always feasible approach (though adding types at the expression level, say, as
in TypeScript should be considered well within bounds).  {\em But\/}, the
concept of studying the idea of an economically rational actor has produced
benefits to the discipline of political economics, and this paper suggests that
implementing and studying the rational programmer will help language designers.

\subsection{How to Turn the Idea of the Rational Programmer into a Methodology} 
\label{sub:methodology}

\input{rational-methodology.tex}

>>>>

This paper dubs Lazarek et. al's idea {\em the rational programmer\/}. Like Stuart
Mill's {\it homo economicus\/}, the rational programmer approximates the
behavior of a software developer who reduces time spent on a task by exploiting the
available information. In the context of gradual typing, the rational programmer
has two pieces of information when an impedance mismatch signals exceptional
behavior: the error message and the state of the program. Hence, the most
rational procedure is to use the former to improve the latter. Specifically,
the rational programmer translates the Wadler-Findler slogan into a debugging
method, searching for the source of the impedance mismatch by adding type
annotations to the untyped parts of the program identified in the error.
If the type checker rejects an annotation that is considered correct, the
rational programmer has found the source of the problem.
Otherwise, the rational programmer concludes that the just-annotated parts are not the problem
and re-runs the program;
the Wadler-Findler slogan says the new version of the program must now blame a different location for the problem, which the rational programmer can check in the same fashion.
Measuring this simulated
programmer's behavior on a large number of debugging scenarios yields data that
is similar to data collected in a human-facing study.

@@

Like {\it homo economicus\/}, the rational programmer is an approximation.
People do not behave in a purely rational manner as economic actors, and they
also do not do so as software developers. The point is not to deny the existence
of ``lucky hunches'' programmers or ``tinkering works'' approaches and so on. It
is also not to claim that equipping entire modules with types represents an
always feasible approach (though adding types at the expression level, say, as
in TypeScript should be considered well within bounds).  {\em But\/}, the
concept of studying the idea of an economically rational actor has produced
benefits to the discipline of political economics, and this paper suggests that
implementing and studying the rational programmer will help language designers.

%% -----------------------------------------------------------------------------
\subsection{How to Turn the Idea of the Rational Programmer into a Methodology} 
\label{sub:methodology}

\input{why-rational-methodology.tex}
